---
title: "Diamonds"
author: "Vijay Sathish"
date: "Sunday, April 15, 2015"
output: html_document
---

## Diamonds dataset  (L3 - 04/15/2015)

```{r}
library(ggplot2)
# diamonds is a built-in data frame in R
data(diamonds)
summary(diamonds)
nobs <- length(row.names(diamonds))
nobs               # 53940; Alternatively, could also use dim(diamonds) to figure this out
length(names(diamonds))          #10 vars
str(diamonds)             # shows three ordered variables

qplot(data = diamonds, x = price,
      color = I('black'), fill = I('#aa88ff'), binwidth = 500) +
  scale_x_continuous(breaks = seq(1000, 19000, 1000), limits = c(0, 19000))

```


### Shape and distribution

```{r}
# It is positively skewed with long right tail
# Mean - 3933; Median - 2401
# Median is a better measure of center or avg. in this case because of the long tail

# Logarthmic converts to bimodal 
qplot(data = diamonds, x = price, 
      color = I('black'), fill = I('#aa88ff')) +
  scale_x_log10()


dl500 <- dim(subset(diamonds, !is.na(diamonds$price) & diamonds$price < 500))
dl500

dl250 <- dim(subset(diamonds, !is.na(diamonds$price) & diamonds$price < 250))
dl250

dg15k <- dim(subset(diamonds, !is.na(diamonds$price) & diamonds$price >= 15000))
dg15k

```

### Ascertain price peak by zooming in
```{r}
# This obviously depends on bin width (frequency around 12k+ for bin width of 500)
# freq. ~160 for binwidth of 1 and occurs at $605
qplot(data = diamonds, x = price,
      color = I('black'), fill = I('#aa88ff'), binwidth = 1) +
  scale_x_continuous(breaks = seq(600, 620, 5), limits = c(600, 620))
#  scale_x_continuous(breaks = seq(1000, 19000, 1000), limits = c(0, 16000))
# + coord_cartesian(ylim = c(10000, 13000))

# Break out prices by cut
qplot(data = diamonds, x = price,
      color = I('black'), fill = I('#aa88ff'), binwidth = 500) +
  scale_x_continuous(breaks = seq(1000, 19000, 4000), limits = c(0, 19000)) +
  facet_wrap(~cut, ncol = 2)


#Break out prices mathematically to get details
by(diamonds$price, diamonds$cut, summary)
# use max if you want absolute accurate value
by(diamonds$price, diamonds$cut, max)

```


### Break free the dimensions using facet_wrap()
```{r}
qplot(data = diamonds, x = price,
      color = I('black'), fill = I('#aa88ff'), binwidth = 500) +
  scale_x_continuous(breaks = seq(1000, 19000, 4000), limits = c(0, 19000)) +
  facet_wrap(~cut, ncol = 2, scales = "free_y")

# price per carat faceted by cut (Remember not to use previous binwidth for log10 scaling)
# Most important reason is to apply transformation to convert to 'normal' like distributions so that linear
# regression can then be applied on it!
by(diamonds$price/diamonds$carat, diamonds$cut, summary)

qplot(data = diamonds, x = price/carat,
      color = I('black'), fill = I('#aa88ff')) + 
  scale_x_log10() +
  facet_wrap(~cut, ncol = 2, scales = "free_y")

```

### Explore summaries and pick most interesting factor for boxplot

```{r}
# Remember that we want variance of price, so that should be y
by(diamonds$price, diamonds$cut, summary)
by(diamonds$price, diamonds$clarity, summary)
by(diamonds$price, diamonds$color, summary)

qplot(data = diamonds, y = price, x = cut,
      color = I('black'), fill = I('#aa88ff'), geom = "boxplot") +
  coord_cartesian(ylim = c(0, 7500))
# Observation - 
# - We see that the median price for 'ideal'  is ~$1800. On, the other hand, its ~$3200 for fair
# - This is obviously telling us that cut alone is not a sufficient indicator of diamond price (ex clarity, color, carat etc)

qplot(data = diamonds, y = price, x = clarity,
      color = I('black'), fill = I('#aa88ff'), geom = "boxplot") +
  coord_cartesian(ylim = c(0, 7000))
# Observation -
# Wider variation in medians (notably that worst clarity has highest median price aka inv. prop)

qplot(data = diamonds, y = price, x = color,
      color = I('black'), fill = I('#aa88ff'), geom = "boxplot") +
  coord_cartesian(ylim = c(0, 9000)) + 
  facet_wrap(~cut, ncol = 3)
# Observation -
# See the same inv. relation (remember D is best and J is worst)

# Observation - After normalizing by carat, the price doesn't seem to vary all that much with color
# Use $color as the fill value to generate box plots with different colors
qplot(data = diamonds, y = price/carat, x = color,
      color = I('black'), fill = color, geom = "boxplot") +
  coord_cartesian(ylim = c(0, 6000))

```

### frequency polygons for carat investigation

```{r}
# Need to set your bin size really really low (0.001) (how accurate to decimal place we want this to be is not mentioned in question)
# Based on question's accuracy, one would assume binwidth=0.1 which is not what question expects though
qplot(data = diamonds, x = carat,
      geom = "freqpoly", binwidth = 0.1) 

qplot(data = diamonds, x = carat,
      geom = "freqpoly", binwidth = 0.001) +
  scale_x_continuous(limits = c(0, 2.5), breaks = seq(0, 2.5, 0.1)) +
  coord_cartesian(ylim = c(0, 3000))

```

## Diamonds dataset  (L4 - 04/19/2015)

### Create a price vs. x scatter plot
```{r}
summary(diamonds$x)
ggplot(data = diamonds, aes(x = x, y = price)) +
  geom_point(alpha = 1/20, color = 'red') +
#  geom_line(stat = median, color = 'black') 
  scale_x_continuous(breaks = seq(0, 11, 1)) +
  scale_y_continuous(breaks = seq(0, 20000, 1000))

```
Response: 
- From the graph, seems like price is a quadratic function of x
- Most of the price is concentrated around 7000

### Correlation prices vs. x, y and z
```{r}
r.px <- with(data = diamonds, cor.test(x, price))
r.px
r.py <- with(data = diamonds, cor.test(y, price))
r.py
r.pz <- with(data = diamonds, cor.test(z, price))
r.pz

```

### Create a price vs. depth scatter plot 
```{r}
summary(diamonds$depth)
ggplot(data = diamonds, aes(x = depth, y = price)) +
  geom_point(alpha = 1/100, color = 'red') +
  scale_x_continuous(limits = c(30, 80), breaks = seq(43, 79, 2))

r.pdepth <- with(data = diamonds, cor.test(depth, price))
r.pdepth

```
Response: As seen from scatterplot, no relation between depth and price


### Price vs. carat scatterplot 
```{r}
# exclude top 1% of price and carat
with(diamonds, quantile(carat, probs = 0.99))
data.q99 <- subset(diamonds, carat <= quantile(carat, probs = 0.99) &
                     price <= quantile(price, probs = 0.99))
ggplot(data = data.q99, aes(x = carat, y = price)) +
  geom_point(alpha = 1/20, color = 'red')

```

### Price vs. volume scatterplot
```{r}
diamonds$volume <- with(diamonds, x*y*z)
ggplot(data = diamonds, aes(x = volume, y = price)) +
  geom_point(alpha = 1/20, color = 'red')


ggplot(data = diamonds, aes(x = volume, y = price)) +
  geom_point(alpha = 1/20, color = 'red') +
  scale_x_log10()

# There are 20 diamonds with a volume 0!
dim(subset(diamonds, volume == 0))  

```
Response: 
- Based on graph, there seem to be extreme outliers in volume
- Majority of volume is bounded by 500, so a log scale will probably give a better picture
- Scaling x axis logarthmically indicates that price is exponentially related to volume

### Correlation b/w price and volume
```{r}
r.vol <- with(subset(diamonds, volume != 0 & volume <= 800), cor.test(volume, price))
r.vol


```

### Insert linear model using geom_smooth() for previous plot excluding outliers as before

```{r}
data.sub <- subset(diamonds, volume != 0 & volume <= 800)
ggplot(data = data.sub, aes(x = volume, y = price)) +
  geom_point(alpha = 1/20, color = 'green') +
  geom_smooth(method = 'lm', color = 'blue') +
  geom_smooth(method = 'lm', formula = y ~ poly(x, 2), color = 'black') +
  geom_smooth(method = 'lm', formula = y ~ poly(x, 3), color = 'black', linetype = 2) +
  coord_cartesian(xlim = c(0, 400), ylim = c(0, 20000))


```
Response: 
- We knew linear model doesn't work well
- scaling x logarithmically showed us this relation already
- We see that cubic fits much better of the three


### Diamonds by clarity (Conditional Means) using dplyr

```{r}
# Name the data frame diamondsByClarity

# The data frame should contain the following
# variables in this order.
#       (1) mean_price
#       (2) median_price
#       (3) min_price
#       (4) max_price
#       (5) n

library('dplyr')
diamondsByClarity <- diamonds %>%
  group_by(clarity) %>%
  summarise(mean_price = mean(as.numeric(price)),
            median_price = median(as.numeric(price)),
            min_price = min(price),
            max_price = max(price),
            n = n()) %>%
  arrange(clarity)

diamondsByClarity

```

### Conditional Means by clarity and color
```{r}
diamonds_by_clarity <- group_by(diamonds, clarity)
diamonds_mp_by_clarity <- summarise(diamonds_by_clarity, mean_price = mean(price))

diamonds_by_color <- group_by(diamonds, color)
diamonds_mp_by_color <- summarise(diamonds_by_color, mean_price = mean(price))

# Plot bar graphs
library('gridExtra')
p1 <- ggplot(data = diamonds_mp_by_clarity, aes(x = clarity, y = mean_price)) +
  geom_bar(stat = 'identity')
p2 <- ggplot(data = diamonds_mp_by_color, aes(x = color, y = mean_price)) +
  geom_bar(stat = 'identity')
grid.arrange(p1, p2, ncol = 1)

```
Response: 
- NOTE: The stat='identity' was required for it to work, otherwise we don't see anything
- With color & clarity, we see the surprising inverse relation with price
- We had earlier noticed this in boxplots as well while faceting by color and clarity separately

## Diamonds dataset  (L5 - 04/20/2015)

### Histogram
```{r}
# Create a histogram of diamond prices. Facet the histogram by diamond color
# and use cut to color the histogram bars


qplot(data = diamonds, x = price, color = 'black', fill = cut, binwidth = 500) +
  facet_wrap(~color, scales = "free_y") 

```

### Scatterplot
```{r}
# Create a scatterplot of diamond price vs. table and color the points by the cut of
# the diamond.

ggplot(data = diamonds, aes(x = table, y = price)) +
  geom_point(aes(color = cut), alpha = 1/5)

```

#### Response: 
Table range for premium cut -> 58-62
Table range for ideal cut -> 53-57

```{r Diamond price vs volume}
# Create a scatterplot of diamond price vs. volume and color the points by
# the clarity of diamonds. Use scale on the y-axis to take the log10 of price. You should also
# omit the top 1% of diamond volumes from the plot.

diamonds$volume <- with(diamonds, x * y * z)
ggplot(data = subset(diamonds, volume <= quantile(x = diamonds$volume, p = 0.99)),
       aes(x = volume, y = log10(price))) +
  geom_point(aes(color = clarity), alpha = 1/1) + 
  scale_color_brewer(type = 'div')
```

```{r}
# Create a scatter plot of the price/carat ratio of diamonds. The variable x should be
# assigned to cut. The points should be colored by diamond color, and the plot should be
# faceted by clarity.

ggplot(data = diamonds, aes(x = cut, y = price/carat)) +
  geom_jitter(aes(color = color), alpha = 1/5) +
  facet_wrap(~clarity)

```













