---
title: "Data Wrangling Basics"
author: "Vijay Sathish"
date: "Friday, April 17, 2015"
output: html_document
---

### Source of info
Notes:
https://s3.amazonaws.com/udacity-hosted-downloads/ud651/DataWranglingWithR.pdf
Data from: http://www.gapminder.org/data/
http://forums.udacity.com/questions/100177970/weekly-working-hours#ud651
Search for 'arms exp' in search bar and download that.
This is very important stuff to convert xlsx forms to data frames

### .xlsx to .csv - After downloading spreadsheet, save into .csv format 
Notes: Only the first tab 'data' will get saved, but thats what we really want. 

### Open up spreadsheet
Notes: When you open it up, you notice that the first row has all the column labels from the xlsx sheet
Hence, when you read in to R, set the header to true. We also set row.names = 1 because first column has country names 
and should get transformed to row names

```{r}
armsRaw <- read.csv("D:/EDA/R_wd/L3/arms_exports.csv", header = T, row.names = 1, check.names = F)
names(armsRaw)
head(armsRaw)
dim(armsRaw)

```

### Transform using tidyr
Notes:
Use gather(<rawdata>, '<col1>', '<col2>', [colRangeLow:colRangeHigh])
Notice that we have 52 columns (1960-2011) and 213 observations, one per country

```{r}
# Transform col1 to col2 for given range
library(tidyr)
armsReshape <- gather(armsRaw, 'year', 'deal_size', 1:52)
head(armsReshape)

```
Response: Notice the column names now

### Almost what we wanted...
Notes: But we managed to not tranform the row names which had country. Trick is to make that a column in original
read by not setting row.names = 1

```{r}
armsRaw2 <- read.csv("D:/EDA/R_wd/L3/arms_exports.csv", header = T, check.names = F)
names(armsRaw2)
head(armsRaw2)
dim(armsRaw2)


```
### Reshape
Notes: Remember that first column is as we want it, so remember to exclude from transform

```{r}
armsReshape2 <- gather(armsRaw2, 'year', 'deal$', 2:53)
head(armsReshape2)
names(armsReshape2)
dim(armsReshape2)

```

### Rename columns and get rid of NAs
```{r}
names(armsReshape2) <- c('country', 'year', 'deal_size')
armsFinal <- subset(armsReshape2, !is.na(deal_size))
head(armsFinal)
dim(armsFinal)
names(armsFinal)

```

***

### Ready to plot

```{r Deals per year}
str(armsFinal)
library(ggplot2)
qplot(data = armsFinal, x = year,
      color = I('black'), fill = I('#aa88ff'),
      xlab = 'year', ylab = '#arms deals per year') +
  scale_x_discrete(breaks = seq(1960, 2011, 5))

```
Response: Both country and year are factors while deal size is not

```{r Median deals per year}
# Convert deal size to logarithmic to better view box plots
qplot(data = armsFinal, y = deal_size, x = year, 
      geom = 'boxplot') +
#  coord_cartesian(ylim = c(0, 500000000)) +
  scale_y_log10()

```
Response: Wide variation in deal sizes over the years. However, we don't see the trend in median increasing deal sizes, so safe to conclude that its somewhat constant

```{r Deals by a country}
# Too many to view in one graph
qplot(data = armsFinal, x = country,
      color = I('black'), fill = I('#aa88ff'),
      xlab = 'country', ylab = '#arms deals per country')  
```

### This would have taken a lot of time! But this really looks like in-depth analysis
Notes: http://forums.udacity.com/questions/100164265/exploring-arms-exports#ud651
Probably would be good to first filter out the top 10 exporters and then analyze only those
What I learned from this exercise is that we basically have too many factors in countries to do useful analysis on full data set





