---
title: "FbBirthdays"
author: "Vijay Sathish"
date: "Friday, April 17, 2015"
output: html_document
---

### Distribution of your friends' birth months and days.
**********************************************************************
How many people share your birthday? Do you know them?
(Reserve time with them or save money to buy them a gift!)
Which month contains the most number of birthdays?
How many birthdays are in each month?
Which day of the year has the most number of birthdays?
Do you have at least 365 friends that have birthdays on everyday
of the year?
**********************************************************************

You will need to do some data munging and additional research to
complete this task. This task won't be easy, and you may encounter some
unexpected challenges along the way. We hope you learn a lot from it though.

You can expect to spend 30 min or more on this task depending if you
use the provided data or obtain your personal data. We also encourage you
to use the lubridate package for working with dates. Read over the documentation
in RStudio and search for examples online if you need help.

You'll need to export your Facebooks friends' birthdays to a csv file.
You may need to create a calendar of your Facebook friends? birthdays
in a program like Outlook or Gmail and then export the calendar as a
csv file.

Once you load the data into R Studio, you can use the strptime() function
to extract the birth months and birth days. We recommend looking up the
documentation for the function and finding examples online.

### Start
Notes: Working with pre-processed list from instructor notes
- Didn't have to go through all the transformations to get to CSV
- No names or year data here, only day and month

```{r Extracting csv to dataframe}
# read.csv() and read.table() achieve same results
raw <- read.csv("D:/EDA/R_wd/L3/birthdaysExample.csv", as.is = T, header = T)
str(raw)
head(raw)

dat.in <- read.table("D:/EDA/R_wd/L3/birthdaysExample.csv", header = T, as.is = T)
str(dat.in)
head(dat.in)

# Use lubridate and this actually becomes a breeze
# http://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html
library('lubridate')
dat.gd <- mdy(dat.in$dates)
str(dat.gd)
head(dat.gd)

# Finally extract out the month, date and years and we are ready to go
dat.df <- data.frame(Value = dat.gd, Year = year(dat.gd), Month = month(dat.gd), Day = day(dat.gd))
str(dat.df)
head(dat.df)
```
Response: Note that read.csv() or read.table() will both work, but important to set as.is = T to prevent from getting it converted into a factors and retain as chr

### Lets see if we can use separate() from tidyr to achieve the same thing!
```{r}
data.untidy <- read.table("D:/EDA/R_wd/L3/birthdaysExample.csv", header = T, as.is = T)
str(data.untidy)
library(tidyr)
data.tidy <- separate(data = data.untidy, col = dates, into = c("Month", "Date", "Year"), sep = "/")
# Conversion to integer from character
data.tidy$Month <- as.integer(data.tidy$Month) 
data.tidy$Date <- as.integer(data.tidy$Date)
data.tidy$Year <- as.integer(data.tidy$Year)
str(data.tidy)
```
Notes: In fact, we can! And this one is actually more intuitive to me than using lubridate()

### Some very basic plots
```{r Basic plots}
library(ggplot2)

qplot(data = dat.df, x = Month,
      color = I('black'), fill = I('#aa88ff'), binwidth = 1, 
      xlab = 'month of the year', ylab = '#birthdays in the month') +
  scale_x_discrete(breaks = c(1:12))

qplot(data = dat.df, x = Day,
      color = I('black'), fill = I('#aa88ff'), binwidth = 1, 
      xlab = 'Day of the year', ylab = '#birthdays on the day') +
  scale_x_discrete(breaks = seq(1, 31, 2))

```