Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes: 04/18/2015

***

### Scatterplots
Notes:

```{r Scatterplots}
pf <- read.csv("D:/EDA/R_wd/L4/pseudo_facebook.tsv", sep = '\t')
str(pf)

library(ggplot2)
qplot(data = pf, x = age, y = friend_count)

```

***

#### What are some things that you notice right away?
Response:
 - Young people (< 30years) seem to have people with extremely large number of friends upto 5000
- Mostly, other age groups have less than 1000 friends.
- Also some spikes for ~68 and again for 100+. Are there really people who are that old? Unlikely

***

### ggplot Syntax
Notes: http://www.ceb-institute.org/bbs/wp-content/uploads/2011/09/handout_ggplot2.pdf
- Limit age to (13, 90) based on min age and based on fact that we don't believe FB has some many users above 90yrs
- Unlike gplot(), with ggplot(), you always have to explicitly specify geom_*()

```{r ggplot Syntax}
summary(pf$age)
ggplot(data = pf, aes(x = age, y = friend_count)) +
  geom_point() +
  xlim(13, 90)

```

***

### Overplotting
Notes: 
- Some of the points, towards the bottom are so closely packed, we can't really tell density
- Introduce alpha component for pixel intensity (alpha = 1/x) means x points needed to make a solid point
- Introduce jitter because age is not a perfect integer

```{r Overplotting}
ggplot(data = pf, aes(x = age, y = friend_count)) +
  geom_point(alpha = 1/20) +
  xlim(13, 90)

ggplot(data = pf, aes(x = age, y = friend_count)) +
  geom_jitter(alpha = 1/20) +
  scale_x_continuous(limits = c(13, 90), breaks = seq(10, 90, 5)) +
  scale_y_continuous(breaks = seq(0, 5000, 200))
  

```

#### What do you notice in the plot?
Response:
- Majority of users below < 400 friends even for young age groups. 
- Still see the spike at age 69 though
***

### Coord_trans()
Notes:
http://docs.ggplot2.org/current/coord_trans.html
- Transform y axis to look at it better using sqrt


```{r Coord_trans()}
ggplot(data = pf, aes(x = age, y = friend_count)) +
  geom_point(alpha = 1/20) +
  xlim(13, 90) +
  coord_trans(y = "sqrt")


```
#### What do you notice?
Response
- Didn't work with geom_jitter(). Had to use geom_point().
- Lines became fainter close to y = 0 indicating that there are very few accounts with close to 0 friends. 

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function with jitter. Create your plot!

Notes:
- Notice that just using geom_jitter() adds jitter to friend count as well.  For accounts with zero users, jitter can make friend_count -ve and sqrt(-ve) will break down

```{r}
# h = 0 specifies that even after y jitter, limit min value of y to 0
ggplot(data = pf, aes(x = age, y = friend_count)) +
  geom_point(alpha = 1/20, position = position_jitter(h = 0)) +
  xlim(13, 90) +
  coord_trans(y = "sqrt")
```

***

### Alpha and Jitter
Explore the relationship between friends_initiated and age
Notes:

```{r Alpha and Jitter}
ggplot(aes(x = age, y = friendships_initiated), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h = 0)) +
  scale_x_continuous(limits = c(13, 90), breaks = seq(10, 90, 5))

summary(pf$friendships_initiated)       # median 46

```
Notes:
- We see that friendships_initiated is lot less than friend_count
- This makes sense because if a user has x friends, we can assume roughly x/2 intiated by user and x/2 are requests to user.
- In fact, we saw median for friend_count was 82 and for friendships_initiated it is 46

***

### Dplyr package
Notes:
http://rstudio-pubs-static.s3.amazonaws.com/11068_8bc42d6df61341b2bed45e9a9a3bf9f4.html
Video: http://www.r-bloggers.com/hadley-wickham-presents-dplyr-at-user-2014/
***

### Conditional Means
Notes: We want to be able to see measures of center variations instead of looking at each point
- Since age goes all the way from 13, 90, its not a good idea to do this using boxplot or histogram for each age. (That would make it 90 different plots)
- So we will transform data to calculate the mean and median of friend counts by age and use that
- Think of dplyr helping in aggregation functions that we see in SQL

```{r Conditional Means}
library('dplyr')
# Does nothing but group observations together by age
pf.by_age <- group_by(pf, age)
head(pf.by_age)

# NOTE that summarise() is different from summary()
pf.fc_by_age <- summarise(pf.by_age, 
                          fc_mean = mean(friend_count), 
                          fc_median = median(friend_count), 
                          n = n())
# for some reason, I already see summarise arranging in order (by accident?)
head(pf.fc_by_age)

pf.fc_by_age <- arrange(pf.fc_by_age, age)
head(pf.fc_by_age)
str(pf.fc_by_age)
```

```{r Data frame pipelining}
# Note that you CANNOT pipeline data tables or databases
pf.fc_by_age <- pf %>%
  group_by(age) %>%
  summarise(fc_mean = mean(friend_count),
            fc_median = median(friend_count),
            n = n()) %>%
  arrange(age)

head(pf.fc_by_age)

```
Response: Note - When pipelining, we omit the data variable because dataframe is assumed to pipeline from previous stage to the next 

### Create your plot!
Notes: 
- Since we have only one data point per mean, we are NOT dealing with frequencies
- This means that histograms, freqplots & boxplots are out of the question
http://docs.ggplot2.org/current/geom_line.html

```{r Conditional Means Plot}
summary(pf.by_age$friend_count)
ggplot(aes(x = age, y = fc_mean), data = pf.fc_by_age) + 
  geom_line() +
  scale_x_continuous(limits = c(13, 90), breaks = seq(10, 90, 5))

```

***

### Overlaying Summaries with Raw Data
Notes:
- Add mean, median, 10% and 90% quantiles on top of scatter plot to get a better sense of ranges
- Wow, some really funky syntax. summary is in quotes, but not the function name mean!

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h = 0), color = 'green') +
  scale_x_continuous(limits = c(13, 90), breaks = seq(10, 90, 5)) +
  coord_trans(y = "sqrt") +
  geom_line(stat = "summary", fun.y = mean) +
  geom_line(stat = "summary", fun.y = median, color = I('#aa88ff')) +
  geom_line(stat = "summary", fun.y = quantile, probs = 0.1, color = I('red'), linetype = 2) +
  geom_line(stat = "summary", fun.y = quantile, probs = 0.9, color = I('red'), linetype = 3)


```

#### What are some of your observations of the plot?
Response:
- Mean is higher than median which means long tailed to right (as we knew before)
- There are quite a bit of outliers even above 90% quartile for younger age groups (ie stddev is high)

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes: 
- Use cor.test to calculate pearson coefficient for age and friend count
- From graph, we see that it is really a bell-curve relation between age & friend count
- So, I am guessing pretty close to 0 or un-correlated

```{r Correlation}
pearson <- cor.test(x = pf$age, y = pf$friend_count)
pearson
r1 <- cor.test(x = pf$age, y = pf$friend_count, method = "kendall", conf.level = 0.95)
r1

```

Look up the documentation for the cor.test function. What's the correlation between age and friend count? Round to three decimal places.

Response:
r = -0.027 
Note how it is almost zero as expected.

***

### Correlation on Subsets
Notes: 
- Do the same thing with data frames using with()
- If we blow away the higher ages (say >65), we can expect a -ve slope line and stronger correlation

```{r Correlation on Subsets}
pearson2 <- with(pf, cor.test(age, friend_count, conf.level = 0.99))
pearson2

p.le_65 <- with(subset(pf, age <= 65), cor.test(age, friend_count, conf.level = 0.99))
p.le_65

```
Response: Notice how we blew away about 11k observations (df went from 99k to 88k)

***

## Create Scatterplots
Notes: 
- Create a scatterplot of likes_received (y) vs. www_likes_received (x)
- Obviously doesn't make sense to add jitter here because likes are discrete
- We zoom in based on summaries. 

```{r}
summary(pf$www_likes_received)
summary(pf$likes_received)
ggplot(aes(y = www_likes_received, x = likes_received), data = pf) + 
  geom_point(alpha = 1/20, color = 'blue')

```
Notes:
- Bulk of data in lower left corner. Get rid of outliers by calculating 95% quantiles for x and y

***

### Strong Correlations
Notes: Also plot the straight line using linear regression

```{r Strong Correlations}
xq <- with(pf, quantile(x = likes_received, probs = 0.95))
xq
yq <- with(pf, quantile(x = www_likes_received, probs = 0.95))
yq
# Use in setting limits
ggplot(aes(y = www_likes_received, x = likes_received), data = pf) + 
  geom_point(alpha = 1/20, color = 'orange') +
  scale_y_continuous(limits = c(0, 227), breaks = seq(0, 227, 20)) +
  scale_x_continuous(limits = c(0, 561), breaks = seq(0, 561, 50)) +
  geom_smooth(method = 'lm', color = 'black')

```
Response: 
- From scatterplot, we expect r to come in very close to 1

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
with(pf, cor.test(likes_received, www_likes_received))

```

Response:

***

### Moira on Correlation
Notes:
- As we already noted, if we use perfectly correlated variables as features for linear regression, it will never converge on a solution. When you use matrices, inverse will not exist. 
- Basically need to identify and get rid of redundant features to get rid of the issue. 

***

```{r More Caution With Correlation}
# install.packages('alr3')
library(alr3)
```

Create your plot!

```{r Temp vs Month}
ggplot(aes(y = Temp, x = Month), data = Mitchell) + 
  geom_point(color = 'blue')

```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.
close to 0 

b. What is the actual correlation of the two variables?
(Round to the thousandths place)


```{r Noisy Scatterplots}
with(Mitchell, cor.test(x = Month, y = Temp))

```

***

### Making Sense of Data
Notes: Bunch up observations by month (ex: month 13 is really january again)

```{r fig.width = 12, fig.height=3}
ggplot(aes(y = Temp, x = Month), data = Mitchell) + 
  geom_point(color = 'blue') +
  scale_x_discrete(limits = c(0, 203), breaks = seq(0, 203, 12))


```

***

### A New Perspective

What do you notice?
Response: Temperature actually varies with month, but not linearly of course . You will notice sinusoidal pattern when you make x axis 5x the y axis in zoomed plot


***

### Understanding Noise: Age to Age Months
Notes: You can do modulo 12 on month to see the same thing more clearly

```{r Understanding Noise: Age to Age Months}

ggplot(aes(y = Temp, x = (Month %% 12)), data = Mitchell) + 
  geom_point(color = 'blue') +
  scale_x_discrete(limits = c(0, 11), breaks = seq(0, 11, 2))

# A multivariate t-test for independence rejects null hypothesis which in this case that they are uncorrelated
# Null hypo is rejected thus proving variables are indeed correlated in some fashion
library('energy')
dcor.ttest(Mitchell$Month, Mitchell$Temp)

```

***

### Calculating Age with Months 
Notes: Current date is Dec 2013

```{r Age with Months Means}
pf$age_with_months <- (2013 - pf$dob_year) + (12 - pf$dob_month)/12
# We could also just do -
# pf$age_with_months <- pf$age + (12 - pf$dob_month)/12
head(pf)

```


### Age with Months Means solution
```{r Programming Assignment}

pf.fc_by_age_months <- pf %>%
  group_by(age_with_months) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age_with_months)

```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}

plot2 <- ggplot(data = subset(pf.fc_by_age_months, age_with_months < 71), aes(x = age_with_months, y = friend_count_mean)) +
  geom_smooth() +
  geom_line() +
  scale_x_continuous(limits = c(13, 71), breaks = seq(13, 71, 5))


```

***

### Smoothing Conditional Means
Notes: Compare against smoother plot 

```{r Smoothing Conditional Means}
plot1 <- ggplot(aes(x = age, y = fc_mean), data = subset(pf.fc_by_age, age < 71)) + 
  geom_line() +
  geom_smooth() +
  scale_x_continuous(limits = c(13, 71), breaks = seq(13, 71, 5))

library(gridExtra)
grid.arrange(plot1, plot2, ncol = 1)

```
Response: This is a trade-off between bias and variance. geom_smooth handles this for you. 
Note however, that geom_smooth() can completely miss out on important info such as the peak at 69.

***

### Which Plot to Choose?
Notes: Choice is based on audience and it may be good to use different plots based on story you are trying to tell

***

### Analyzing Two Variables
Reflection:
- scatter plots with geom_point()
- adding jitter with geom_jitter() 
- using alpha for transparency to mitigate overplotting
- Pipelines using dply with group_by(), summarise() and arrange() to calculate conditional means df
- Overlaying mean, median and quantiles on top of scatter plots
- calculating pearson's coefficient with and doing a correlation test with cor.test()
- Use correlation to not pick variables that are strongly correlated in feature set as this can 
lead to logistic/linear regression breaking down
- multivariate ttest for independence between two variables using dcor.ttest()
- conditional means plots with geom_line() after calculating a df with central measures using dply
- noise and smoothing and which plots to choose

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

