Lesson 3
========================================================

***

### What to Do First?
Notes: Download the psuedo fb data set

***

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
setwd("D:/EDA/R_wd/L3")
pf <- read.csv("pseudo_facebook.tsv", sep = '\t')
names(pf)

```

***

### Histogram of Users' Birthdays
Notes: 
http://flowingdata.com/2014/02/27/how-to-read-histograms-and-use-them-in-r/
- ggplot2 is easier to use than the plotting functions provided in 'base' pkg
- qplot is really a histogram plot
- scale_x_discrete() is used to separate out the bars to make it more viewable

```{r Histogram of Users\' Birthdays}
# install.packages('ggplot2')
library(ggplot2)
qplot(data = pf, x = dob_day) +
  scale_x_discrete(breaks=1:31)
```

***

#### What are some things that you notice about this histogram?
Response:
- We find that the graph is roughly uniform distributed for data. 
- The bar for 31 is smaller but expected since only 7 months in a year have 31 days.
- The one bar that stands out is for 1st of the month. But this probably indicates we 
don't have data for all subjects and the model defaulted to 1 in case of N/A.

***

### Moira's Investigation
Notes:
http://hci.stanford.edu/publications/2013/invisibleaudience/invisibleaudience.pdf
***

### Estimating Your Audience Size
Notes:

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response:
Ski trip to Tahoe with selfies on Squaw and Kirkwood.

#### How many of your friends do you think saw that post?
Response: 30 friends or about 10%

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response: 

***

### Perceived Audience Size
Notes:

***
### Faceting
Notes:
http://www.cookbook-r.com/Graphs/Facets_(ggplot2)/
- facet by dob_month; ncol is just how many graphs per column
- This will basically break up histogram into 12, one per month

```{r Faceting}
qplot(data = pf, x = dob_day) +
  scale_x_discrete(breaks=1:31) +
  facet_wrap(~dob_month, ncol = 3)      
  

```

#### Let’s take another look at our plot. What stands out to you here?
Response:
- All the monthly histograms also look uniform as expected
- Here, we see the spike for Jan 1st which must be the DOB default for FB

***

### Be Skeptical - Outliers and Anomalies
Notes:
- Important thing to note is which anamolies are actually real and which ones are due to some issue with data
- Top coded data - Sometimes, data points whose values are above an upper bound are censored
- In general, it is important to think about which category an anamoly falls into, so that it can be adjusted or discarded before analysis

***

### Moira's Outlier
Notes:
#### Which case do you think applies to Moira’s outlier?
Response:
Bad data about an extreme case

***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
qplot(data = pf, x = friend_count)
```

#### How is this plot similar to Moira's first plot?
Response:
- Data skewed to right meaning most people have 0-100 friends which falls into one bucket and a long tail

***

### Limiting the Axes
Notes:
http://docs.ggplot2.org/current/scale_continuous.html
- Since bulk of our data is for friend_count < 1000, we limit our histogram using xlim

```{r Limiting the Axes}
qplot(data = pf, x = friend_count, xlim = c(0, 1000))
# equivalent syntax
# qplot(data = pf, x = friend_count) + 
#   scale_x_continuous(limits = c(0, 1000))

```

### Exploring with Bin Width
Notes:

***

### Adjusting the Bin Width
Notes:

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender, ncol = 2)
```

***

### Omitting NA Values
Notes:
- using na.omit(pf) instead will get rid of all observations that have NA in at least one variable
- For this particular situation, might not be the best option
- limits gives range of x axis
- breaks states how often the ticks appear on the axis

```{r Omitting NA Values}
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender, ncol = 2)

# qplot(x = friend_count, data = na.omit(pf), binwidth = 25) +
#   scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
#   facet_wrap(~gender, ncol = 2)
```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
# split friend_count by gender into two buckets and provide summary for each bucket
24

```

#### Who on average has more friends: men or women?
Response: Women

#### What's the difference between the median friend count for women and men?
Response:

#### Why would the median be a better measure than the mean?
Response:
Because we have a long tail to the right, the mean would tend to over-estimate the average. (more robust)
***

### Tenure
Notes:
- You can substitute actual color names by hex RGB values #RRGGBB
- color --> outline color; I() --> as is
- fill --> what the bars get filled with

```{r Tenure}
qplot(data = pf, x = tenure, color = I('black'), fill = I('blue'), binwidth = 30)
# qplot(data = pf, x = tenure, color = I('black'), fill = I('#0000ff'), binwidth = 30)

```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
qplot(data = pf, x = tenure/365, color = I('black'), fill = I('blue'), binwidth = 0.25) +
  scale_x_continuous(limits = c(0, 7), breaks = seq(0, 7, 1))
```

***

### Labeling Plots
Notes:

```{r Labeling Plots}
qplot(data = pf, x = tenure/365,
      xlab = 'Num. years in fb',
      ylab = 'Num users per bin',
      color = I('black'), fill = I('blue'), binwidth = 0.25) +
  scale_x_continuous(limits = c(0, 7), breaks = seq(0, 7, 1))

```

***

### User Ages
Notes:

```{r User Ages}
qplot(data = na.omit(pf), x = age, binwidth = 1,
      xlab = 'Age in years of FB users', ylab = 'Num users per bin',
      color = I('black'), fill = I('#aa88ff')) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(10, 100, 10)) +
  facet_wrap(~gender)

```

#### What do you notice?
Response:
- Noticed that majority of users in the 20-30 age group
- Also, there is spike in interest for 50-65 year olds. Grandma's?

***

### The Spread of Memes
Notes:

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
Notes:
```{r}
summary(pf$friend_count)

summary(log10(pf$friend_count + 1))    # +1 is because log 0 will be -inf

```


***

### Add a Scaling Layer
Notes:
http://lightonphiri.org/blog/ggplot2-multiple-plots-in-one-graph-using-gridextra
- ggplot is also quite similar to qplot except wrapping x and y args within aes
- aes stands for aesthetic
- note that scale_x_log10() and scale_x_sqrt() can be used w/ qplot() as well

```{r Add a Scaling Layer}

library(gridExtra)
p1 <- qplot(data = pf, x = friend_count, binwidth = 20,
      xlab = 'user friend count', ylab = 'frequency of users per bin',
      color = I('black'), fill = I('#aa88ff')) +
  scale_x_continuous(limits = c(0, 3000), breaks = seq(0,3000, 200))
# Amazing how log10 plot transformed into a normal distribution!
p2 <- qplot(data = pf, x = log10(friend_count+1), binwidth = 0.1,
      xlab = 'log10(user friend count + 1)', ylab = 'frequency of users per bin',
      color = I('black'), fill = I('#aa88ff')) +
  scale_x_continuous(limits = c(0, 4), breaks = seq(0, 4, 0.5))

p3 <- qplot(data = pf, x = sqrt(friend_count), binwidth = 1,
      xlab = 'sqrt(user friend count)', ylab = 'frequency of users per bin',
      color = I('black'), fill = I('#aa88ff')) +
  scale_x_continuous(limits = c(0, 70), breaks = seq(10, 70, 10))

grid.arrange(p1, p2, p3, ncol = 1)

# ggplot syntax to do the same
# p4 <- ggplot(data = pf, aes(x = friend_count)) + geom_histogram()
# p5 <- p3 + scale_x_log10()          # notice that this is still labeling in x units
# p6 <- p3 + scale_x_sqrt()
# grid.arrange(p3, p4, p5, ncol = 1)


```

***


### Frequency Polygons
Notes: 
- Uses the geom arg for qplot
- notice how y is set to change to proportion rather than absolute to compare male and female
```{r Frequency Polygons}
qplot(data = subset(pf, !is.na(gender)), x = friend_count, y = ..count../sum(..count..), 
      binwidth = 20, geom = 'freqpoly',
      xlab = 'user friend count', ylab = 'frequency of users per bin',
      color = gender) +
  scale_x_continuous(limits = c(0, 3000), breaks = seq(0,3000, 200)) 
# + scale_y_continuous(limits = c(0, 0.005))   # to zoom into crossover of females over males
# happens around 800 friend_count

```

***

### Likes on the Web
Notes:

```{r Likes on the Web}
qplot(data = subset(pf, !is.na(gender)), x = www_likes, 
      geom = 'freqpoly',
      xlab = 'user friend count', ylab = 'frequency of users per bin',
      color = gender) + 
  scale_x_log10()

```

### Likes on the web solution
Notes: Using this, it seems that women have about 2x the number of likes

```{r}
by(pf$www_likes, pf$gender, sum)
```

***

### Box Plots
Notes:
- Remember that y is always the frequency that you want to count
- x is always what you want to divide by

```{r Box Plots}
qplot(data = subset(pf, !is.na(gender)),
      x = gender, y = friend_count,
      geom = 'boxplot')

```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.
Notes: 
- Although, we may think the solution is to subset the data to get rid of outliers above 1000,
that will change the nature of the boxplot itself which is WRONG!
- Note that scale_y_continuous(limits = c(0, 1000)) and ylim variable have the same problem as well!
- Correct way is to use coord_cartesian() to only adjust the axis
```{r}
# qplot(data = subset(pf, !is.na(gender) & friend_count <= 1000),
#      x = gender, y = friend_count,
#      geom = 'boxplot')
qplot(data = subset(pf, !is.na(gender)),
      x = gender, y = friend_count,
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 1000))
  
```

***

### Box Plots, Quartiles, and Friendships
Notes:
- Check that the summary matches our box plots after using coord_cartesian()
```{r Box Plots, Quartiles, and Friendships}
by(pf$friend_count, pf$gender, summary)

```

#### On average, who initiated more friendships in our sample: men or women?
Response: Women
#### Write about some ways that you can verify your answer.
Response:
```{r Friend Requests by Gender}
by(pf$friendships_initiated, pf$gender, summary)
by(pf$friendships_initiated, pf$gender, sum)    # although total is higher for men
```

Response:
Both mean and median higher for women than men
***

### Getting Logical
Notes: 
- Important to change to factors to see summary for $mobile_checkins
```{r Getting Logical}
summary(pf$mobile_likes)
summary(pf$mobile_likes > 0)

mobile_checkins <- NA
pf$mobile_checkins <- ifelse(pf$mobile_likes > 0, 1, 0)
pf$mobile_checkins <- factor(pf$mobile_checkins)      
summary(pf$mobile_checkins)

num <- sum(pf$mobile_checkins == 1)
den <- length((pf$mobile_checkins))
num * 100 / den

```

Response:
***
Almost 65% of the users have used mobile checkins.
This means that continuing to develop FB app for mobile platform is very important. 

### Analyzing One Variable
Reflection:

***
