Lesson 6 - 04/21/2015
========================================================

### Chris Saden's analysis of the same diamond data set
https://s3.amazonaws.com/udacity-hosted-downloads/ud651/diamondsExample.html

### Scatterplot Review
Notes: Real dataset from 2008

```{r Scatterplot Review}

library('ggplot2')
data(diamonds)
ggplot(data = diamonds, aes(x = carat, y = price), 
       xlim = c(0, quantile(carat, probs = 0.99)),
       ylim = c(0, quantile(price, probs = 0.99))) +
  geom_smooth(method = 'lm', color = 'blue', linetype = 2) +
  geom_point(color = 'red')

```

Response:
- Price = fexp(carat)
- But we also see variance increase as the carat size increases

***

### ggpairs Function
Notes: Scatter Plot matrix
- quant-quant --> displays r and scatterplots
- qual-quant --> displays histogram and boxplots
- qual- qual --> displays histogram and boxplots again

```{r ggpairs Function}
# install these if necessary
# install.packages('GGally')
# install.packages('scales')
# install.packages('memisc')
# install.packages('lattice')
# install.packages('MASS')
# install.packages('car')
# install.packages('reshape')
# install.packages('plyr')

# load the ggplot graphics package and the others
library(ggplot2)
library(GGally)
library(scales)
library(memisc)

# sample 10,000 diamonds from the data set
set.seed(20022012)
# Easier to follow syntax - some error, debug later
# sample_price <- sample(x = diamonds$price, size = 10000)
# diamond_samp <- subset(diamonds, sample_price %in% price)
# head(diamond_samp)
diamond_samp <- diamonds[sample(1:length(diamonds$price), 10000), ]
ggpairs(diamond_samp, params = c(shape = I('.'), outlier.shape = I('.')))
ggsave("D:/EDA/R_wd/Project/sp_matrix.pdf")
```

What are some things you notice in the ggpairs output?
Response:
- x, y and z are almost perfectly correlated to each other (means diamonds always come in same shapes)
- price is also quite strongly correlated to x, y, z (ie volume of diamond)
- carat is strongly correlated to price and x, y, z (which we expected)
- depth shows no impact on price (reject feature for price prediction)
- We already saw the weird inverse relation of color and clarity to price (against intuition)
- Nothing substantial stands out wrt. to cut, but we cannot eliminate it just yet
- Hypothesize that since weight is a function of volume, price may be related to cube of carat

***

### The Demand of Diamonds
Notes: 
- Maybe the case that in the smaller diamonds, there is less affluent buyers, so this is price sensitive. This makes the variation of price for sub-carat diamonds not show much variation
- For larger sizes, there is not much demand, and demand is from affluent people. So, possibly much wider variation in prices for weight > 1 carat

- May be instrumental to transform the price variable using log
http://www.r-statistics.com/2013/05/log-transformations-for-skewed-and-wide-distributions-from-practical-data-science-with-r/

```{r The Demand of Diamonds}
library(gridExtra)

plot1 <- qplot(data = subset(diamonds, price > 0) , x= price, color = 'black', fill = I('#ff88aa'), binwidth = 100) + 
  ggtitle('Price')

# Note, in log plot, you have to pick bin width after the transformation
plot2 <- qplot(data = subset(diamonds, price > 0), x = price, color = 'black', fill = I('#aa88ff'), binwidth = 0.01) +
  scale_x_log10() +
  ggtitle('Price (log10)') 

grid.arrange(plot1, plot2, ncol = 1)
```

Notes: - 
- Price distribution is +vely skewed with long tail. Hence, log10 transformation should help convert to near normal
- Note that for applying LR, we only require inputs to be approx normal, not the the output itself
- Money is also multiplicative - year on year inflation, path dependent (rich become richer) etc
- For monetary variables, which are highly skewed and have orders of magnitude variance, it is useful to apply log transformation
- This also helps data models which don't do well when there is orders of magnitude difference
Post, transformation, we see it is bimodal, but closer to normal, so we will apply LR on log10(y) instead of y

***

### Connecting Demand and Price Distributions
Notes: 
- First peak for price senstive customers
- Second peak is for affluent customers. Beyond a point, they also don't think its worth it apparently

***

### Scatterplot Transformation
Notes: Transform price using log10 due to it being reasons above
```{r Scatterplot Transformation}
ggplot(data = diamonds, aes(x = carat, y = price), 
       xlim = c(0, quantile(carat, probs = 0.99)),
       ylim = c(0, quantile(price, probs = 0.99))) +
  geom_point(color = 'red', alpha = 1/5) +
  scale_y_log10()

```

### Some useful functions
```{r Basic Functions - log(x), log(x^2), log(x^3) }
input <- seq(0, 10, 0.1)
ylin <- input
ysq <- input^2
ycube <- input^3
func.df <- data.frame(input, ylin, ysq, ycube)
str(func.df)
p1 <- ggplot(data = func.df, aes(x = input, y = ylin)) +
  geom_point(color = 'blue', linetype = 2) +
  scale_y_log10()
p2 <- ggplot(data = func.df, aes(x = input, y = ysq)) +
  geom_point(color = 'red', linetype = 3) +
  scale_y_log10()
p3 <- ggplot(data = func.df, aes(x = input, y = ycube)) +
  geom_point(color = 'green') + 
  scale_y_log10()
grid.arrange(p1, p2, p3, ncol = 1)

ylin <- input
ysqr <- input^(1/2)
ycuber <- input^(1/3)
func.df <- data.frame(input, ylin, ysq, ycube)
str(func.df)
p4 <- ggplot(data = func.df, aes(x = input, y = ylin)) +
  geom_point(color = 'blue', linetype = 2)
p5 <- ggplot(data = func.df, aes(x = input, y = ysqr)) +
  geom_point(color = 'red', linetype = 3) +
  scale_y_log10()
p6 <- ggplot(data = func.df, aes(x = input, y = ycuber)) +
  geom_point(color = 'green') + 
  scale_y_log10()
grid.arrange(p4, p5, p6, ncol = 1)

```


### Create a new function to transform the carat variable
Notes:
https://www.youtube.com/watch?v=Z1wB1rHAYzQ&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP
http://blog.ggplot2.org/post/25938265813/defining-a-new-transformation-for-ggplot2-scales

Note: Write a general function called 'cuberoot_trans()' that returns the cube root of input param
```{r cuberoot transformation}
cuberoot_trans = function() trans_new('cuberoot', transform = function(x) x^(1/3),
                                      inverse = function(x) x^(3))
```

#### Use the cuberoot_trans function
```{r Use cuberoot_trans}
ggplot(aes(carat, price), data = diamonds) + 
  geom_point(color = 'red', alpha = 1/5) +
#  scale_x_continuous(trans = cuberoot_trans()) +
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat')
```

Response: 
- Viola, it looks linear now 
- Note, the cube root hint also comes from the strong correlation of each dimension to price


***

### Overplotting Revisited

```{r Sort and Head Tables}
head(sort(table(diamonds$price), decreasing = T))     # sorts by frequency
head(sort(table(diamonds$carat), decreasing = T))

```

Response: Shows you can really end up overplotting with the number of points at the same carat/price level

```{r Overplotting Revisited}
ggplot(aes(carat, price), data = diamonds) + 
  geom_jitter(alpha = 0.5, size = 0.75) + 
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat')
```

***

### Other Qualitative Factors
Notes:

***

### Price vs. Carat and Clarity
Notes:
http://www.bluenile.com/education/diamonds/cut
http://www.bluenile.com/education/diamonds/color
http://www.bluenile.com/education/diamonds/clarity

http://docs.ggplot2.org/current/scale_brewer.html
http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#palettes-color-brewer
http://www.cookbook-r.com/Graphs/Legends_(ggplot2)/

```{r Price vs. Carat and Clarity}
# install and load the RColorBrewer package
# install.packages('RColorBrewer')
# library(RColorBrewer)

# legend just tells you what to do while printing legend (overrides alpha and size so it is visible better)
# Reverse is because we want clarity displayed in descending order
ggplot(aes(x = carat, y = price), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter', aes(color = clarity)) +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Clarity', reverse = T,
    override.aes = list(alpha = 1, size = 4))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
    breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
    breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Color')
```


***

### Clarity and Price
Response:
Based on the plot, we can see that when the clarity increases, we see the price leaning towards the higher end of the band. 

***

### Price vs. Carat and Cut

Alter the code below.
```{r Price vs. Carat and Cut}
ggplot(aes(x = carat, y = price, color = cut), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Cut', reverse = T,
                                          override.aes = list(alpha = 1, size = 4))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Cut')
```

***

### Cut and Price
Response:
Based on plot, cuts are dispersed along Y axis, so no clear relation to cut and price

***

### Price vs. Carat and Color

```{r Price vs. Carat and Color}
ggplot(aes(x = carat, y = price, color = color), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Color', reverse = F,
                                          override.aes = list(alpha = 1, size = 4))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Color')
```

***

### Color and Price
Response: Based on this, we see that color plays a role similar to clarity, but the boundaries are less distinct. This is most likely because D to J are all still excellent colors with color barely detectable except in I-J

***

### Linear Models in R
Notes:
use lm(output ~ input_func) to perform linear regression


Response:
- Use lm(log(price) ~ carat^(1/3)) based on our plot

***

### Building the Linear Model
Notes: http://data.princeton.edu/R/linearModels.html
Interpreting Regression Coefficients in R - 
  http://www.r-bloggers.com/interpreting-regression-coefficient-in-r/?utm_source=feedburner&utm_medium=email&utm_campaign=Feed%3A+RBloggers+%28R+bloggers%29
Interpreting Regression Coefficients on the Analysis Factor blog
  http://www.theanalysisfactor.com/interpreting-regression-coefficients/
Fitting and Interpreting Linear Models by yhat 
  http://blog.yhathq.com/posts/r-lm-summary.html
Another Explanation of Factor Coefficients in Linear Models on Stats StackExchange
  http://stats.stackexchange.com/a/24256

```{r Building the Linear Model}
m1 <- lm(I(log(price)) ~ I(carat^(1/3)), data = diamonds)
m2 <- update(m1, ~ . + carat)
m3 <- update(m2, ~ . + cut)
m4 <- update(m3, ~ . + color)
m5 <- update(m4, ~ . + clarity)
mtable(m1, m2, m3, m4, m5)
```

Response
- Notice how adding cut to our model does not help explain much of the variance in the price of diamonds. This fits with out exploration earlier.
- Also see how R-squared approches 1 showing that our model is very close to actual price!

***

### Model Problems
http://www.bain.com/publications/articles/global-diamond-report-2013.aspx
Video Notes:

Response:
- Diamond prices as a whole can dip or rise based on supply-demand. For example, 2008 recession resulted in steep drop in prices due to fall in demand
- We also see that lower carat diamonds show more variation with time
- Data we have is based on single year ie 2008. So, it doesn't take time into account 
- So can't simply 

***

### A Bigger, Better Data Set
Notes: The code used to obtain the data is available here:
https://github.com/solomonm/diamonds-data

```{r A Bigger, Better Data Set}
# This is not required since we have downloaded dataset directly
# install.package('bitops')
# install.packages('RCurl')
library('bitops')
library('RCurl')

# Downloaded data
# diamondsurl = getBinaryURL("https://raw.github.com/solomonm/diamonds-data/master/BigDiamonds.Rda")
# load(rawConnection(diamondsurl))
load("BigDiamonds.Rda")
head(diamondsbig, 100)
```

## Building a Model Using the Big Diamonds Data Set
Notes: Only consider diamonds that are GIA certified and price < 10000

```{r Building a Model Using the Big Diamonds Data Set}
diamondsBigSample <- subset(diamondsbig, cert == "GIA" & !is.na(price) & price <= 10000) 
m1 <- lm(I(log10(price)) ~ I(carat^(1/3)), data = diamondsBigSample)
m2 <- update(m1, ~ . + carat)
m3 <- update(m2, ~ . + cut)
m4 <- update(m3, ~ . + color)
m5 <- update(m4, ~ . + clarity)
mtable(m1, m2, m3, m4, m5)


```

Note: Same model does quite well even for the big data set

***

## Price Prediction using the Linear Model

Example Diamond from BlueNile:
Round 1.00 Very Good I VS1 $5,601

```{r}
#Be sure you’ve loaded the library memisc and have m5 saved as an object in your workspace.
thisDiamond <- data.frame(carat = 1.00, cut = "V.Good",
                         color = "I", clarity="VS1")
modelEstimate <- predict(m5, newdata = thisDiamond,
                        interval="prediction", level = .95)

priceEst <- 10^modelEstimate
priceEst

 
```

Evaluate how well the model predicts the BlueNile diamond's price. Think about the fitted point estimate as well as the 95% CI.

***

## Final Thoughts
Notes:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

